### 2021-07-08 rad-icfp-2021/complexity.tex
rewriting complexity section

In this section, we introduce a simple cost model on our language and show 
that our reverse-mode transformation satisfies a version of the cheap gradient principle (Theorem \ref{thm:complexity}).

@@

By inspection in Figure~\ref{fig:direct_diff_macro}, 
we see that there is a linear usage of  each continuation $Y$ 
and that it is always applied to almost the identity. 
More precisely, we have applications of the form !fun (x1,$\ldots$,xn) -> Y(e1,$\ldots$,en)! 
where the !ei! are !xi! except for at most $k$ (independent of $n$) terms.
In fact we have $k=2$, except for the !map2! and !reduce! cases which require a more detailed analysis (see Section~\ref{sec:complexity}).
We can first use inlining as given in Figure~\ref{fig:optim}.
Using the invariant above, most of the !e!$_{i}$ below are variables. 
Without loss of generality, assume the only terms !e!$_{i}$ which are not variables are !e!$_{n-1}$ and !e!$_{n}$.
We can then use forward substitution on the other !e!$_{i}$.
In brief:

@@

Though this transformation after the inlining above already has the right complexity, 
its purity allows us to make the most out of generic optimizations.  
A list of common optimizations that are useful for this language can be found in Figure~\ref{fig:optim}.

@@

For simplicity, we will not address any parallelism in our cost model. 
So the cost of !map (x.e)! on an array of size $n$ will be $n*\cost$(!e!).
The cost function extends compositionally to the whole language in the obvious way.
For instance we have !$\cost$(let x=e$_1$ in e$_2$)!= !$\cost$(e$_1$)+$\cost$(e$_2$)!.
TODO: for let binding and same in map, do I need to add some MOVES?

@@

\begin{theorem}
    \label{thm:complexity}
    Given a term !$\Gamma \vdash$ e: $\reals$! such that $\NestA$(!e!)$\leq$ $p$.
    Denote by $G$ the term $\grad_\Gamma$!e! after inlining and forward substitution.
    TODO: be more precise with "inlining and forward substitution".
    Then !$\cost$(G) $\leq$ $4*2^{p}*\cost$(e)!.
\end{theorem}

@@

In Figure~\ref{fig:optim} we present a of optimizations. 
These optimizations are all obviously valid according to the semantic model. 
As our language is purely functional, we can use these extra optimizations aggressively. 
A lot of simplifications come from the ring structure of the reals, lifted to tuples and arrays.
After these optimizations, the output program is essentially a sequence of let bindings 
and resembles SSA-code.

>>>>

In this section, we introduce a simple cost model for our language and show 
that, after partial evaluation, our reverse-mode transformation satisfies a version of the cheap gradient principle (Theorem \ref{thm:complexity}).

@@

\begin{lemma}
    Let $\Gamma \vdash$!e: A! be a term in Source. 
    Every variable !Y! in $\directD{\rho}{\Gamma}{Y}$(!e!) which is not of ground type is used exactly once.
\end{lemma}

\begin{proof}
    By induction on the judgment of !e!. We inspect each rule in Figure~\ref{fig:direct_diff_macro}.
\end{proof}

From the previous lemma, we see that there is a linear usage of  each continuation $Y$. 
This is one key property which has been ensured by $\UNFSymbol$.
We also note that, apart from the !map2! case, this continuation variable is always applied to almost the identity. 
More precisely, we have applications of the form !fun (x1,$\ldots$,xn) -> Y(e1,$\ldots$,en)! 
where the !ei! are !xi! except for at most $k$ (independent of $n$) terms.
In fact we have $k=2$, except for the !map2! and !reduce!.

We can first use inlining, the first optimization rule given in Figure~\ref{fig:optim}.
Using the invariant above, most of the !e!$_{i}$ below are variables. 
Without loss of generality, assume the only terms !e!$_{i}$ which are not variables are !e!$_{n-1}$ and !e!$_{n}$.
We can then use forward substitution, the third optimization rule in Figure~\ref{fig:optim}, on the other !e!$_{i}$.
In brief:

@@

We call this optimization step partial evaluation in the rest of the complexity section. 
We have the following result.

\begin{lemma}
    \label{lem:noHO}
    Let $\Gamma \vdash$!e: A! be a term in Source. 
    After the partial evaluation step, $\directD{\rho}{\Gamma}{Y}$(!e!) does not have lambda abstractions.
    Its only variable (bound and free) which does not have a ground type is !Y!.
\end{lemma}

\begin{proof}
    By induction on the judgment of !e!. 
    We inspect each rule in Figure~\ref{fig:direct_diff_macro} 
    and note that the partial evaluation step precisely allows us to conclude the inductive step.
\end{proof}

@@

For simplicity, we will not address any parallelism in our cost model. 
So the cost of !map (x.e)! on an array of size $n$ will be $n*\cost$(!e!).
Following Lemma \ref{lem:noHO}, we do not need our cost model to deal with higher-order variables and lambda abstractions.
We can thus restrict our attention to the subset of the Target language which does not contain lambdas, applications or variables which are not of ground type.
The cost function extends compositionally to the restricted Target language. 
It is given in Figure~\ref{fig:costmodel}.

\input{figures/costmodel}

@@

We can now phrase our main complexity theorem.

\begin{theorem}
    \label{thm:complexity}
    Given a term !$\Gamma \vdash$ e: $\reals$! such that $\NestA$(!e!)$\leq$ $p$.
    Denote by $G$ the term $\grad_\Gamma$!e! after the partial evaluation step from Section~\ref{sub:Partial evaluation and optimization}.
    Then !$\cost$(G) $\leq$ $4*2^{p}*\cost$(e)!.
\end{theorem}

@@

In Figure~\ref{fig:optim} we present a of optimizations. 
These optimizations are all obviously valid according to the semantic model. 
As our language is purely functional, we can use these extra optimizations aggressively. 
A lot of simplifications come from the ring structure of the reals, lifted to tuples and arrays.
After these optimizations, the output program is essentially a sequence of let-bindings 
and resembles SSA-code \cite{cytron1989efficient}.

Even though our reverse-mode transformation has the right complexity after the partial evaluation step, 
the constant factor is of huge importance in practice. 
The purity of our transformation allows us to make the most out of generic optimizations.  
A list of common optimizations that are useful for this language can be found in Figure~\ref{fig:optim}.
In addition, hand-crafted efficient derivatives and more optimizations can easily be added to our language.
